A Sound and Complete Deep Embedding of Arbitrary Normal Modal Logics in Coq

### 2024-01-27 paper/COLA/paper.tex
Little changes.

\subsection{An overview of modal logic.} Modal logics are designed for reasoning
about truth across various --- abstract --- worlds. In such logics, a
proposition may be true in some world, but false in another one. The systems of
modal logic considered in this work extend traditional propositional logic
with two operators on propositions: $\Box$ (box) and $\Diamond$ (diamond).
The syntax for modal logic formulas is defined by the following context free
grammar (where
$\mathcal{P} = \{A,B,C,...\}$ is the countable set of propositional variables):
\[
  \begin{array}{lcl}
    \varphi & ::= & \mathcal{P}\,|\,\neg\varphi\,|\,\varphi\land\varphi\,|\,\varphi\lor\varphi\,|\,\varphi\to\varphi\,|\,\Box\varphi\,|\,\Diamond\varphi
  \end{array}  
\]
Intuitively, the semantics of the operator $\Box$ is similar to an universal
quantifier, since it expresses \emph{necessity}: $\Box\,\varphi$ is true in the
current world iff $\varphi$ is true in every world accessible from the current
one. Similarly, the semantics of operator
$\Diamond$ is that a formula $\Diamond\varphi$ is true in the current
world iff $\varphi$ is \emph{possible}, i.e. true in at least some world
accessible from the current world. We the remaining of the paper, we also write
$\phi_1\leftrightarrow\phi_2$ to mean $\phi_1\to\phi_2\land\phi_2\to\phi_1$.

@@

\begin{figure}[H]
\begin{enumerate}
  % \item $\mathcal{M}\,;\,w\not\models \bot$ and $\mathcal{M}\,;\,w\models \top$;
  \item $\mathcal{M}\,;\,w\models A$ iff $w \in \mathcal{V}(A)$;
  \item $\mathcal{M}\,;\,w\models \neg\varphi$ iff $\mathcal{M}\,;\,w\not\models\varphi$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \land \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ and $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \lor \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \to \varphi_2$ iff
    $\mathcal{M}\,;\,w\not\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models\Box\varphi$ iff $\forall y \in W. w R y \to \mathcal{M}\,;\,y \models\varphi$;
  \item $\mathcal{M}\,;\,w\models\Diamond\varphi$ iff $\exists y \in W. w R y \land \mathcal{M}\,;\,y \models\varphi$;
  \end{enumerate}
  \caption{Kripke semantics}
  \label{fig:kripke}
\end{figure}
We say that a formula $\varphi$ is satisfiable in a model $\mathcal{M}$,
denoted by $\mathcal{M}\models \varphi$, if $\forall w \in W. M\,;w\,\models \varphi$ holds.
Similarly, $\varphi$ is valid in a frame $\mathcal{F}$, written $\mathcal{F}\models\varphi$, if
it holds on all models $\mathcal{M} = \langle \mathcal{F},\mathcal{V}\rangle$, i.e.
for any $\mathcal{M}$, we have that $\mathcal{M}\models\varphi$. A formula
is valid, written $\models \varphi$, if it is valid in any frame $\mathcal{F}$.

Using the semantics, we can define the concept of \emph{entailment}: given a set
of formulas $\Gamma$ and a formula $\varphi$, we say that $\Gamma$ entails
$\varphi$, written $\Gamma\models\varphi$, iff $\forall \mathcal{M} . (\forall \varphi' \in\Gamma .
\mathcal{M} \models \varphi') \to \mathcal{M} \models \varphi$.

In this work, we consider Hilbert deductive systems for modal logic.
Essentially, the considered system adds axioms for the modalities
to the set of axioms for classical propositional logic.
As examples of modalities axioms we mention \textbf{K}, which express that $\Box$ distributes
over implication, and possibility which states that $\Diamond$ distributes
over disjunction.
\[
  \begin{array}{ll}
    \textbf{K} & \Box(\varphi_1 \to \varphi_2) \to (\Box \varphi_1 \to \Box \varphi_2)\\
    \textbf{Possibility} & \Diamond (\varphi_1 \lor \varphi_2) \to
                           (\Diamond \varphi_1 \lor \Diamond \varphi_2)\\
  \end{array}
\]
Apart from the axioms, we also need \emph{modus ponens} and necessitation rules.

@@

A central aspect in the study of modal logics is the so-called \emph{frame
  correspondence}, which is a relationship between the
accessibility relation and the truth of certain formulas in frames containing
it. As an example, consider a frame $\mathcal{F} = \langle W,R
\rangle$ in which the relation $R$ is reflexive: $\forall w \in W. wRw$, i.e.
every world is accessible from itself. In such frame when $\Box A$ holds in
a world $w$, $w$ itself is among the accessible worlds in which $A$ must be true.
In this way, if the accessibility relation $R$ is reflexive, then whatever world
$w$ and formula $\varphi$ we take, the formula $\Box \varphi \to \varphi$ will
be true there. Examples of properties that characterize certain formula schemas are
presented in the table below.

\begin{figure}[H]
\begin{tabular}{|c|c|}
  \hline
  Property of $R$                          & Formula schema \\ \hline
  reflexive: $\forall w.wRw$               & $\Box A \to A$ \\ 
  serial: $\forall w\exists w'. wRw'$      & $\Box A \to \Diamond A$ \\
  symmetric: $\forall w\,w'.wRw' \to w'Rw$ & $A \to \Box\Diamond A$ \\
  \hline
\end{tabular}
\centering
\end{figure}
There are other properties of interest for a relation $R$ like being
transitive, functional, euclidean, and so on. More about the correspondence
of frames and deductive systems properties can be found elsewhere~\cite{Chellas,blackburn}.

@@

Basically, a Hilbert-style deductive system allows the use of four
possibilities in a proof: 1) the use of a hypothesis; 2) the use of an axiom;
3) the use of the modus ponens rule or 4) the use of the generalization (necessity)
rule. The inductive type \mintinline{coq}{deduction} encodes these possibilities
as its constructors. The \mintinline{coq}{deduction} type is parameterized by
the set of axioms that form a system for some modal logic. In this way, our definition can be
used by several systems alike.
\begin{minted}{coq}
Inductive deduction (A: axiom -> Prop):
                          theory -> formula -> Prop :=
  (* Premise. *)
  | Prem: forall (t: theory)
                 (f: formula),
        t f -> deduction A t f
  (* Axiom. *)
  | Ax: forall (t: theory)
               (a: axiom)
               (f: formula),
        A a -> instantiate a = f -> deduction A t f
  (* Modus Ponens. *)
  | Mp: forall (t: theory)
               (f g: formula)
               (d1: deduction A t (Implies f g))
               (d2: deduction A t f),
        deduction A t g
  (* Generalization. *)
  | Nec: forall (t: theory)
                (f: formula)
                (d1: deduction A t f),
         deduction A t ([] f).
\end{minted}
The constructors \mintinline{coq}{Prem}, \mintinline{coq}{Mp} and \mintinline{coq}{Nec}
represent rules for hypothesis, modus ponens and generalization, respectively.
Before discussing the \mintinline{coq}{Ax} constructor, we need to discuss how axioms are
represented in our library. The inductive type \mintinline{coq}{axiom} specifies
the different axioms supported by our library.
\begin{minted}{coq}
Inductive axiom: Set :=
  | ax1  : formula -> formula -> axiom
  | ax2  : formula -> formula -> formula  -> axiom
  (* some code omitted *)
  | axK  : formula -> formula -> axiom
  | axPos: formula -> formula -> axiom
  | axT  : formula -> axiom
  | axB  : formula -> axiom
  | axK4 : formula -> axiom
  | axD  : formula -> axiom
  | axK5 : formula -> axiom.
\end{minted}
Each constructor of type \mintinline{coq}{axiom} represents an axiom scheme and
its type stores the subformulas which compose the axiom. The type has
17 constructors: the first 10 encode axioms for propositional logic
and, the last 7 are for different formulas which are taken as axioms by common
systems of modal logic. Axioms are built using the function \mintinline{coq}{instantiate},
which receives an axiom and builds a corresponding formula by pattern matching as
follows:
\begin{minted}{coq}
Definition instantiate (a: axiom): formula :=
  match a with
  | ax1    p   q
     => [! p -> (q -> p) !]
  | ax2    p   q   r
     => [! (p -> (q -> r)) -> ((p -> q) -> (p -> r)) !]
  | ax3    p   q
     => [! (~ q -> ~ p) -> (p -> q) !]
  | axK    p   q
     => [! [] (p -> q) -> ([] p -> [] q) !]
  | axPos  p   q
     => [! <> (p \/ q) -> (<> p \/ <> q) !]
  (* some code omitted *)
  end.
\end{minted}
Notation \mintinline{coq}{[! p !]} is used to write modal logic formulas using a
syntax closer to mathematical practice using Coq. This is necessary to avoid
parsing ambiguities, since some symbols are also used by Coq's syntax.
Modalities $\Box$ and $\Diamond$ are denoted by \mintinline{coq}{[]}
and \mintinline{coq}{<>}, respectively.

@@

In order to use the deductive system to
check the provability of formulas, we need to specify the axioms that are
included in the system. Our approach is to define predicates
which denote each system. The simplest set of axioms is usually referred as
\textbf{K} and it includes the axiom $\Box(A \to B) \to (\Box A \to \Box B)$
and all the axioms for propositional logic. We represent system \textbf{K}
by the following inductive predicate:
\begin{minted}{coq}
Inductive K: axiom -> Prop :=
  | K_ax1: forall p q, K (ax1 p q)
  | K_ax2: forall p q r, K (ax2 p q r)
  (* some code omitted *)
  | K_axK: forall p q, K (axK p q)
\end{minted}
Type \mintinline{coq}{K} is a predicate over axioms: it has one constructor for
each of the valid axioms which form the system. Given a value
of \mintinline{coq}{axiom}, a value of type \mintinline{coq}{K a} denotes a
proof that a given axiom is part of the system \textbf{K}. This is the main
trick used by the constructor \mintinline{coq}{Ax}, of type
\mintinline{coq}{deduction}, to generically represent axioms for
different systems of modal logics uniformly. In order to understand how it
works, let us review its type (below, remember that \mintinline{coq}{A: axiom -> Prop}
denotes a system of axioms for modal logics):
\begin{minted}{coq}
Ax: forall (t: theory)
           (a: axiom)
           (f: formula),
           A a -> instantiate a = f -> deduction A t f
\end{minted}
Note that the fourth parameter of constructor \mintinline{coq}{Ax} is a value
of type \mintinline{coq}{A a}, which is only inhabited if the predicate
\mintinline{coq}{A} holds for the axiom \mintinline{coq}{a}. Through this
evidence, is possible to create the corresponding axiom formula instance by
using function \mintinline{coq}{instantiate}. In this way, we are able to
represent all systems of axioms for modal logics using a single type
for its axioms.

@@

The proof proceeds by induction on the derivation of $\Gamma \vdash\varphi$ and
it has a case for each of the 11 axioms of the Hilbert system for propositional
logics, one for the $K$ axiom and cases for using hypothesis and the rules modus
ponens. In order to modularize the proof, we defined a lemma for each case.
As example, the following is the statement for modus ponens case.
\begin{minted}{coq}
Lemma Modus_Ponens_soundness:
  forall M w p q, (M ' w ||- p) /\ (M ' w ||- [! p -> q !]) -> (M ' w ||- q).
\end{minted}
All these lemmas are direct consequences from the definitions of the Kripke
semantics and entailment. We note that system K is used as it doesn't impose
any restrictions on the frame.

@@

In this section, we describe our formalization of the completeness result for the
Hilbert proof system with respect to the Kripke semantics. The proof consists of two
basic steps: 1) the proof of the existence of maximal consistent sets and; 2) the
construction of a cannonical model from the maximal consistent set. We describe these
steps separately.

\paragraph{Maximal consistent sets} The first step on the completeness proof is the construction of
a \emph{maximal consistent sets}, $\Delta$, of formulas for $\Gamma$ and $\varphi$.  The $\Delta$ set
is defined as follows:

@@

Which states that it is not possible to deduce \verb|p| and \verb|~ p|, for any formula \verb|p|.
In order to define maximal consistent sets, we use the following representation of a theory
subset predicate:

@@

The completeness proof rely on the fact that the set of modal logic formulas is countable. We model
this fact by two functions \mintinline{coq}{encode :: formula -> nat} and \mintinline{coq}{decode :: nat -> option formula}
which satisfies the following property: $\forall \varphi . \texttt{decode(encode }\varphi\texttt{)} = \varphi$.

@@

The first constructor states that when we do not insert a formula in a theory,
we do not change it. The second constructor includes a formula in a theory if
adding that formula keeps the theory consistent and the final constructor considers
the situation when the inclusion of a formula \verb|p| makes the theory inconsistent.
In this case, we add the negation of \verb|p| to the theory.

@@

In order to produce the $\Delta_{i}$, we need to include the $i$-th formula (built using the function \mintinline{coq}{decode})
in the theory $\Delta_{i - 1}$. Finally, the maximal set is defined as:
\begin{minted}{coq}
Definition Max : theory := UnionOf Delta.
Definition UnionOf (A : nat -> theory) : theory := fun t => exists i, A i t.
\end{minted}
Function \mintinline{coq}{UnionOf} is a immediate encoding the indexed set union operator using Coq. An important property
of this construction of the maximal set is that it only increases the number of elements. This fact is proved by
the followed lemma, which has an immediate proof using the previous definitions.
\begin{minted}{coq}
Lemma subset_of_insert: forall D p, Subset D (Insert p D).
\end{minted}
Next, we show that the set \mintinline{coq}{Max} is such that either \verb|p| or \verb|~ p| is an element of \verb|Max|.
\begin{minted}{coq}
Lemma max_is_maximal: forall p, Max [! p !] \/ Max [! ~p !].
\end{minted}
The next step in the proof is showing that \mintinline{coq}{Max} is consistent. In order to prove this fact,
we use some lemmas. First, we show, by case analysis on formulas, that inserting an formula into a theory
preserves consistency:
\begin{minted}{coq}
Lemma insert_preserves_consistency: forall D p, Consistent A D -> Consistent A (Insert p D).
\end{minted}
and using the previous lemma, we prove that operation \mintinline{coq}{Delta n} preserves consistency, by induction on \verb|n|.
\begin{minted}{coq}
Lemma delta_is_consistent: Consistent A G -> forall n, Consistent A (Delta n).
\end{minted}
Next, the proof that \mintinline{coq}{Max} is consistent is a corollary of the previously stated results.
\begin{minted}{coq}
Corollary max_is_consistent: Consistent A G -> Consistent A Max.
\end{minted}
The final piece is the proof of the Lindenbaum's theorem, which states that any consistent theory has
a maximal consistent set. The proof of this fact is immediate from the previous lemmas.
\begin{minted}{coq}
Theorem lindebaum : Consistent A G -> exists D, Consistent A D /\ Maximal D /\ Subset G D.
\end{minted}
This concludes our formalization of the existence of the maximal consistent sets for modal logics formula.

@@

In order to state L\"ob's theorem we need a way to encode modal
fixed points. Formally, let $\varphi(X)$ be a formula with only one propositional variable
$X$. The modal fixed point of $\varphi$ is then a formula $\Psi$ such that such that
$\Psi \leftrightarrow \varphi(\Box \Psi)$.
We represent the existence of modal fixed points for all one-variable formulas by a definition \mintinline{coq}{fixed_point},
which is parameterized by a system of modal logic, \mintinline{coq}{S}, and a
theory, \mintinline{coq}{g}, and encodes the logical equivalence demonstrating
that \mintinline{coq}{p} is a modal fixed point for \mintinline{coq}{f}.
\begin{minted}{coq}
  Definition fixed_point S g: Prop :=
    forall f, exists p,
    (S; g |-- [! (p <-> f ([]p)) !]).
\end{minted}

@@

\begin{minted}{coq}
  Theorem Lob:
    forall A, subset K4 A /\ fixed_point A nil ->
    forall P, (A; nil |-- [! []P -> P !]) ->
              (A; nil |-- [! P !]).
\end{minted}

\noindent The proof starts by assuming an arbitrary formula $P$, the existence of modal fixed
points and that $\Box P \to P$ is provable with axioms A. These deduction steps
are performed by the following tactic:

\begin{minted}{coq}
  intros A [I FP] P H1.
\end{minted}

\noindent Next, from the existence of modal fixed points for every formula,
we know that there is a formula $\Psi$ such that $\Psi \leftrightarrow (\Box
\Psi \to P)$. The following tactic performs this step of reasoning.

@@

\noindent Now, we are able to deduce that $\Psi \to (\Box \Psi\to P)$ from the previous
step result as follows:

\begin{minted}{coq}
  assert (A; nil |-- [! psi -> []psi -> P !]) as H3.
  apply modal_ax5 in H2; auto with modal.
\end{minted}

\noindent The \mintinline{coq}{assert} tactic allows the definition of
auxiliary proof goals. In the previous code piece, we specified
$\vdash \Psi \to (\Box \Psi \to P)$ as a new goal which
generates a new proof hypothesis named \mintinline{coq}{H3}. The
deduction of this fact is immediate from K4 axioms, of which A is a superset.
We use some basic automation to move on.

From $\vdash \Psi \to (\Box \Psi \to P)$ and the necessitation
rule, we can deduce $\vdash \Box (\Psi \to (\Box \Psi \to P))$
using the following tactics:
\begin{minted}{coq}
  assert (A; nil |-- [! [](psi -> []psi -> P) !])
    as H4.
  apply Nec; auto.
\end{minted}

\noindent Now, using the axiom \textbf{K} we can deduce $\vdash \Box \Psi \to
\Box (\Box \Psi \to P)$. Again, we deduce this fact using the
\mintinline{coq}{assert} tactic.

>>>>

\subsection{An overview of modal logic.} Modal logics are designed for reasoning
about truth across various --- abstract --- worlds. In such logics, a
proposition may be true in some world, but false in another one. The systems of
modal logic considered in this work extend traditional propositional logic
with two operators on propositions: $\Box$ (box) and $\Diamond$ (diamond).
The syntax for modal logic formulas is defined by the following context free
grammar (where
$\mathcal{P} = \{A,B,C,...\}$ is the countable set of propositional variables):
\[
  \begin{array}{lcl}
    \varphi & ::= & \mathcal{P}\,|\,\neg\varphi\,|\,\varphi\land\varphi\,|\,\varphi\lor\varphi\,|\,\varphi\to\varphi\,|\,\Box\varphi\,|\,\Diamond\varphi
  \end{array}  
\]
Intuitively, the semantics of the operator $\Box$ is similar to an universal
quantifier, since it expresses \emph{necessity}: $\Box\,\varphi$ is true in the
current world iff $\varphi$ is true in every world accessible from the current
one. Similarly, the semantics of operator
$\Diamond$ is that a formula $\Diamond\varphi$ is true in the current
world iff $\varphi$ is \emph{possible}, i.e. true in at least some world
accessible from the current world. In the remaining of the paper, we also write
$\phi_1\leftrightarrow\phi_2$ to mean $\phi_1\to\phi_2\land\phi_2\to\phi_1$.

@@

\begin{figure}[H]
\begin{enumerate}
  % \item $\mathcal{M}\,;\,w\not\models \bot$ and $\mathcal{M}\,;\,w\models \top$;
  \item $\mathcal{M}\,;\,w\models A$ iff $w \in \mathcal{V}(A)$;
  \item $\mathcal{M}\,;\,w\models \neg\varphi$ iff $\mathcal{M}\,;\,w\not\models\varphi$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \land \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ and $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \lor \varphi_2$ iff
    $\mathcal{M}\,;\,w\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models \varphi_1 \to \varphi_2$ iff
    $\mathcal{M}\,;\,w\not\models\varphi_1$ or $\mathcal{M}\,;\,w\models\varphi_2$;
  \item $\mathcal{M}\,;\,w\models\Box\varphi$ iff $\forall y \in W. (w R y \to \mathcal{M}\,;\,y \models\varphi)$;
  \item $\mathcal{M}\,;\,w\models\Diamond\varphi$ iff $\exists y \in W. (w R y \land \mathcal{M}\,;\,y \models\varphi)$;
  \end{enumerate}
  \caption{Kripke semantics}
  \label{fig:kripke}
\end{figure}
We say that a formula $\varphi$ is satisfiable in a model $\mathcal{M}$,
denoted by $\mathcal{M}\models \varphi$, if $\forall w \in W. (M\,;w\,\models \varphi)$ holds.
Similarly, $\varphi$ is valid in a frame $\mathcal{F}$, written $\mathcal{F}\models\varphi$, if
it holds on all models $\mathcal{M} = \langle \mathcal{F},\mathcal{V}\rangle$, i.e.
for any $\mathcal{M}$, we have that $\mathcal{M}\models\varphi$. A formula
is valid, written $\models \varphi$, if it is valid in any frame $\mathcal{F}$.

Using the semantics, we can define the concept of \emph{entailment}: given a set
of formulas $\Gamma$ and a formula $\varphi$, we say that $\Gamma$ entails
$\varphi$, written $\Gamma\models\varphi$, iff $\forall \mathcal{M} . ((\forall \varphi' \in\Gamma .
\mathcal{M} \models \varphi') \to \mathcal{M} \models \varphi)$.

In this work, we consider Hilbert deductive systems for modal logic.
Essentially, the considered system adds axioms for the modalities
to the set of axioms for classical propositional logic.
As examples of modalities axioms we mention K, which express that $\Box$ distributes
over implication, and possibility which states that $\Diamond$ distributes
over disjunction.
\[
  \begin{array}{ll}
    K & \Box(\varphi_1 \to \varphi_2) \to (\Box \varphi_1 \to \Box \varphi_2)\\
    Possibility & \Diamond (\varphi_1 \lor \varphi_2) \to
                           (\Diamond \varphi_1 \lor \Diamond \varphi_2)\\
  \end{array}
\]
Apart from the axioms, we also need \emph{modus ponens} and necessitation rules.

@@

A central aspect in the study of modal logics is the so-called \emph{frame
  correspondence}, which is a relationship between the
accessibility relation and the truth of certain formulas in frames containing
it. As an example, consider a frame $\mathcal{F} = \langle W,R
\rangle$ in which the relation $R$ is reflexive: $\forall w \in W. (wRw)$, i.e.
every world is accessible from itself. In such frame when $\Box A$ holds in
a world $w$, $w$ itself is among the accessible worlds in which $A$ must be true.
In this way, if the accessibility relation $R$ is reflexive, then whatever world
$w$ and formula $\varphi$ we take, the formula $\Box \varphi \to \varphi$ will
be true there. Examples of properties that characterize certain formula schemas are
presented in the table below.

\begin{figure}[H]
\begin{tabular}{|c|c|}
  \hline
  Property of $R$                          & Formula schema \\ \hline
  reflexive: $\forall w.(wRw)$               & $\Box A \to A$ \\
  serial: $\forall w\exists w'. (wRw')$      & $\Box A \to \Diamond A$ \\
  symmetric: $\forall w\,w'.(wRw' \to w'Rw)$ & $A \to \Box\Diamond A$ \\
  \hline
\end{tabular}
\centering
\end{figure}
There are other properties of interest for a relation $R$ like being
transitive, functional, euclidean, and so on. More about the correspondence
of frames and deductive systems properties can be found elsewhere~\cite{Chellas,blackburn}.

@@

Basically, a Hilbert-style deductive system allows the use of four
possibilities in a proof: 1) the use of a hypothesis; 2) the use of an instantiation of an axiom;
3) the use of the modus ponens rule or 4) the use of the generalization (necessity)
rule. The inductive type \mintinline{coq}{deduction} encodes these possibilities
as its constructors. The \mintinline{coq}{deduction} type is parameterized by
the set of axioms that form a system for some modal logic. In this way, our definition can be
used by several systems alike.
\begin{minted}{coq}
Inductive deduction (A: axiom -> Prop) : theory -> formula -> Prop :=
  (* Premise. *)
  | Prem: forall (t: theory)
                 (f: formula),
        t f -> deduction A t f
  (* Axiom. *)
  | Ax: forall (t: theory)
               (a: axiom)
               (f: formula),
        A a -> instantiate a = f -> deduction A t f
  (* Modus Ponens. *)
  | Mp: forall (t: theory)
               (f g: formula)
               (d1: deduction A t (Implies f g))
               (d2: deduction A t f),
        deduction A t g
  (* Generalization. *)
  | Nec: forall (t: theory)
                (f: formula)
                (d1: deduction A t f),
         deduction A t ([] f).
\end{minted}
The constructors \mintinline{coq}{Prem}, \mintinline{coq}{Mp} and \mintinline{coq}{Nec}
represent rules for hypothesis, modus ponens and generalization, respectively.
Before discussing the \mintinline{coq}{Ax} constructor, we need to discuss how axioms are
represented in our library. The inductive type \mintinline{coq}{axiom} specifies
the different axioms supported by our library.
\begin{minted}{coq}
Inductive axiom: Set :=
  | ax1  : formula -> formula -> axiom
  | ax2  : formula -> formula -> formula  -> axiom
  (* some code omitted *)
  | axK  : formula -> formula -> axiom
  | axPos: formula -> formula -> axiom
  | axT  : formula -> axiom
  | axB  : formula -> axiom
  | axK4 : formula -> axiom
  | axD  : formula -> axiom
  | axK5 : formula -> axiom.
\end{minted}
Each constructor of type \mintinline{coq}{axiom} represents an axiom scheme and
its type stores the subformulas which compose the axiom. The type has
17 constructors: the first 10 encode axioms for propositional logic
and, the last 7 are for different formulas which are taken as axioms by common
systems of modal logic. Axioms are built using the function \mintinline{coq}{instantiate},
which receives an axiom and builds a corresponding formula by pattern matching as
follows:
\begin{minted}{coq}
Definition instantiate (a: axiom): formula :=
  match a with
  | ax1    p   q
     => [! p -> (q -> p) !]
  | ax2    p   q   r
     => [! (p -> (q -> r)) -> ((p -> q) -> (p -> r)) !]
  | ax3    p   q
     => [! (~ q -> ~ p) -> (p -> q) !]
  | axK    p   q
     => [! [] (p -> q) -> ([] p -> [] q) !]
  | axPos  p   q
     => [! <> (p \/ q) -> (<> p \/ <> q) !]
  (* some code omitted *)
  end.
\end{minted}
Notation \mintinline{coq}{[! p !]} is used to write modal logic formulas using a
syntax closer to mathematical practice using Coq. This is necessary to avoid
parsing ambiguities, since some symbols are also used by Coq's syntax.
Modalities $\Box$ and $\Diamond$ are denoted by \mintinline{coq}{[]}
and \mintinline{coq}{<>}, respectively.

@@

In order to use the deductive system to
check the provability of formulas, we need to specify the axioms that are
included in the system. Our approach is to define predicates
which denote each system. The simplest set of axioms is usually referred as
K and it includes the axiom $\Box(A \to B) \to (\Box A \to \Box B)$
and all the axioms for propositional logic. We represent system K
by the following inductive predicate:
\begin{minted}{coq}
Inductive K: axiom -> Prop :=
  | K_ax1: forall p q, K (ax1 p q)
  | K_ax2: forall p q r, K (ax2 p q r)
  (* some code omitted *)
  | K_axK: forall p q, K (axK p q)
\end{minted}
Type \mintinline{coq}{K} is a predicate over axioms: it has one constructor for
each of the valid axioms which form the system. Given a value
of \mintinline{coq}{axiom}, a value of type \mintinline{coq}{K a} denotes a
proof that a given axiom is part of the system K. This is the main
trick used by the constructor \mintinline{coq}{Ax}, of type
\mintinline{coq}{deduction}, to generically represent axioms for
different systems of modal logics uniformly. In order to understand how it
works, let us review its type (below, remember that \mintinline{coq}{A: axiom -> Prop}
denotes a system of axioms for modal logics):
\begin{minted}{coq}
Ax: forall (t: theory)
           (a: axiom)
           (f: formula),
           A a -> instantiate a = f -> deduction A t f
\end{minted}
Note that the fourth parameter of constructor \mintinline{coq}{Ax} is a value
of type \mintinline{coq}{A a}, which is only inhabited if the predicate
\mintinline{coq}{A} holds for the axiom \mintinline{coq}{a}. Through this
evidence, is possible to create the corresponding axiom formula instance by
using function \mintinline{coq}{instantiate}. In this way, we are able to
represent all systems of axioms for modal logics using a single type
for its axioms.

@@

The proof proceeds by induction on the derivation of $\Gamma \vdash\varphi$ and
it has a case for each of the axioms of the Hilbert system for propositional
logics, one for the $K$ axiom and cases for using hypothesis, the rules modus
ponens and necessitation. In order to modularize the proof, we defined a lemma for each case.
As example, the following is the statement for modus ponens case.
\begin{minted}{coq}
Lemma Modus_Ponens_soundness:
  forall M w p q, (M ' w ||- p) /\ (M ' w ||- [! p -> q !]) -> (M ' w ||- q).
\end{minted}
All these lemmas are direct consequences from the definitions of the Kripke
semantics and entailment. We note that system K is used as it doesn't impose
any restrictions on the frame.

@@

In this section, we describe our formalization of the completeness result for the
Hilbert proof system with respect to the Kripke semantics.
It is worth mentioning that there are different notions of completeness and
the literature is not always clear about the difference of those.

{\color{red} TORRENS: Escrever definições de completude}

The proof consists of two
basic steps: 1) the proof of the existence of maximal consistent sets and; 2) the
construction of a cannonical model from the maximal consistent set. We describe these steps separately.

\paragraph{Maximal consistent sets} The first step on the completeness proof is the construction of a \emph{maximal consistent set}, $\Delta$, of formulas for $\Gamma$ and $\varphi$.  The $\Delta$ set
is defined as follows:

@@

Which states that it is not possible to deduce \mintinline{coq}{p} and \mintinline{coq}{~p}, for any formula \mintinline{coq}{p}.
In order to define maximal consistent sets, we use the following representation of a theory
subset predicate:

@@

The completeness proof rely on the fact that the set of modal logic formulas is countable. We model
this fact by two functions \mintinline{coq}{encode :: formula -> nat} and \mintinline{coq}{decode :: nat -> option formula}
which satisfies the following property: $\forall \varphi . (\texttt{decode(encode }\varphi\texttt{)} = \varphi)$.

@@

The first constructor states that when we do not insert a formula in a theory,
we do not change it. The second constructor includes a formula in a theory if
adding that formula keeps the theory consistent and the final constructor considers
the situation when the inclusion of a formula \verb|p| makes the theory inconsistent.
In this case, we add the negation of \verb|p| to the theory. Notation \mintinline{coq}{Extend p D} denotes the union of theory \mintinline{coq}{D} with the
singleton set containing \mintinline{coq}{p}.

@@

In order to produce the $\Delta_{i}$, we need to include the $i$-th formula (built using the function \mintinline{coq}{decode})
in the theory $\Delta_{i - 1}$. Finally, the maximal set is defined as:
\begin{minted}{coq}
Definition Max : theory := UnionOf Delta.
Definition UnionOf (A : nat -> theory) : theory := fun t => exists i, A i t.
\end{minted}
Function \mintinline{coq}{UnionOf} is a immediate encoding the indexed set union operator using Coq. An important property
of this construction of the maximal set is that it only increases the number of elements. This fact is proved by
the followed lemma, which has an immediate proof using the previous definitions.
\begin{minted}{coq}
Lemma subset_of_insert: forall D p, Subset D (Insert p D).
\end{minted}
Next, we show that the set \mintinline{coq}{Max} is such that either \verb|p| or \verb|~p| is an element of \verb|Max|.
\begin{minted}{coq}
Lemma max_is_maximal: forall p, Max [! p !] \/ Max [! ~p !].
\end{minted}
The next step in the proof is showing that \mintinline{coq}{Max} is consistent. In order to prove this fact,
we use some lemmas. First, we show, by case analysis on formulas, that inserting an formula into a theory
preserves consistency:
\begin{minted}{coq}
Lemma insert_preserves_consistency: forall D p, Consistent A D -> Consistent A (Insert p D).
\end{minted}
and using the previous lemma, we prove that operation \mintinline{coq}{Delta n} preserves consistency, by induction on \verb|n|.
\begin{minted}{coq}
Lemma delta_is_consistent: Consistent A G -> forall n, Consistent A (Delta n).
\end{minted}
Next, the proof that \mintinline{coq}{Max} is consistent is a corollary of the previously stated results.
\begin{minted}{coq}
Corollary max_is_consistent: Consistent A G -> Consistent A Max.
\end{minted}
The final piece is the proof of the Lindenbaum's theorem, which states that any consistent theory has
a maximal consistent set. The proof of this fact is immediate from the previous lemmas.
\begin{minted}{coq}
Theorem lindenbaum : Consistent A G -> exists D, Consistent A D /\ Maximal D /\ Subset G D.
\end{minted}
This concludes our formalization of the existence of the maximal consistent set for modal logic theories.

@@

In order to state L\"ob's theorem we need a way to encode modal
fixed points. Formally, let $\varphi(X)$ be a formula with only one propositional variable
$X$. The modal fixed point of $\varphi$ is then a formula $\Psi$ such that
$\Psi \leftrightarrow \varphi(\Box \Psi)$.
We represent the existence of modal fixed points for all one-variable formulas by a definition \mintinline{coq}{fixed_point},
which is parameterized by a system of modal logic, \mintinline{coq}{S}, and a
theory, \mintinline{coq}{G}, and encodes the logical equivalence demonstrating
that \mintinline{coq}{p} is a modal fixed point for \mintinline{coq}{f}.
\begin{minted}{coq}
  Definition fixed_point A G: Prop :=
    forall f, exists p,
    (A; G |-- [! (p <-> f ([]p)) !]).
\end{minted}

@@

\begin{minted}{coq}
  Theorem Lob:
    forall A G, subset K4 A /\ fixed_point A G ->
    forall p, (A; G |-- [! []p -> p !]) ->
              (A; G |-- [! p !]).
\end{minted}

\noindent The proof starts by assuming an arbitrary formula $p$, the existence of modal fixed
points and that $\Box p \to p$ is provable with axioms A. These deduction steps
are performed by the following tactic:

\begin{minted}{coq}
  intros A G [I FP] P H1.
\end{minted}

\noindent Next, from the existence of modal fixed points for every formula,
we know that there is a formula $\Psi$ such that $\Psi \leftrightarrow (\Box
\Psi \to p)$. The following tactic performs this step of reasoning.

@@

\noindent Now, we are able to deduce that $\Psi \to (\Box \Psi\to p)$ from the previous
step result as follows:

\begin{minted}{coq}
  assert (A; G |-- [! psi -> []psi -> P !]) as H3.
  apply modal_ax5 in H2; auto with modal.
\end{minted}

\noindent The \mintinline{coq}{assert} tactic allows the definition of
auxiliary proof goals. In the previous code piece, we specified
$\vdash \Psi \to (\Box \Psi \to p)$ as a new goal which
generates a new proof hypothesis named \mintinline{coq}{H3}. The
deduction of this fact is immediate from K4 axioms, of which A is a superset.
We use some basic automation to move on.

From $\vdash \Psi \to (\Box \Psi \to p)$ and the necessitation
rule, we can deduce $\vdash \Box (\Psi \to (\Box \Psi \to p))$
using the following tactics:
\begin{minted}{coq}
  assert (A; G |-- [! [](psi -> []psi -> p) !])
    as H4.
  apply Nec; auto.
\end{minted}

\noindent Now, using the axiom K we can deduce $\vdash \Box \Psi \to
\Box (\Box \Psi \to p)$. Again, we deduce this fact using the
\mintinline{coq}{assert} tactic.
